A significance test in forward stepwise regression

### 2014-05-01 paper/paper.tex
Reworked power simulations, began comparison with BIC

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some approaches for this problem can be posed as optimization
criteria which stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
(\todo describe construction of $\beta$ after finding $A$)
where we have written $\{ \beta_s : s = 1, \ldots, steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). Stopping rules based on classical test
statistics have also been used, so it is natural to consider using
the new test statistics of \cite{significance:lasso} or
\cite{tests:adaptive}
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section~\ref{sec:simulations}.

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a small segment of the
compressed sensing literature \citep{donoho:pursuit, cai:wang:omp}
with results stating that forward stepwise (often referred to in that
literature as Orthogonal Matching Pursuit or OMP) can exactly select
the true model
under some stringent conditions involving quantities like the sparsity
of the true model and the coherence of the design matrix. The
coherence $\mu(X)$ of a matrix $X$ with columns scaled to have unit
2-norm is defined as

\begin{equation}
  \mu := \mu(X) = \max_{i \neq j} \{ | \innerp{ X_i }{ X_j } | \}
\end{equation}

Recalling that $k$ is the sparsity of $\beta$, typical results in this
literature say that if $k < (1/\mu + 1)/2$ and the nonzero
coefficients of $\beta$ are sufficiently large then forward stepwise
recovers $\beta$ with high probability. The coherence condition is
necessary to guarantee exact recovery \citep{cai:wang:xu:sharp} in the
sense that it is possible to construct counterexamples with
$k = (1/\mu + 1)/2$. We refer the reader to the literature for details.
For our purposes the
conditions required to guarantee exact recovery are usually too
stringent. Simulations show empirically that forward
stepwise can work well even when it is not working perfectly, and that
it does so under a wide range of conditions.

\begin{figure}
\begin{center}
\subfigure[Ternary design]{
\label{fig:fsternary}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/ternary_p2n_n100_lower1p4_upper1p4.pdf}}
\hspace{-15pt}
\subfigure[Gaussian design with correlation]{
\label{fig:fsgausscorr}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows the results of the simulation
  using design matrices with i.i.d. ternary entries taking values $0,
  \pm 1$ with equal probability. The left panel shows the results
  with design matrices of standard Gaussian entries with
  equi-correlation of 0.2. }
\label{fig:fwdstepsim}
\end{center}
\end{figure}

\jonathan{Is average power the same as TPR (or, more correctly, TPP)?}
\josh{Yes, maybe I should have used that terminology instead?}

For various sparsity levels $k$, Algorithm~\ref{algo:fs} was run on
various data sets with \textit{steps} equal to $k$ and the ``average
power'' was calculated as the number of true variables in the active
set after $k$ steps divided by $k$:
$$
\text{Power}_k = \frac{\# \left\{g: g \in A_k, \beta_g \neq 0\right\}}{k}.
$$ (\todo Make this consistent with
$\gamma$ in simulations section)
Nonzero coefficients had magnitude
in a range of multiples of $\gamma := \sqrt{2\log(G)}$, e.g. in $[1.4
\gamma, 1.6 \gamma]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. After computing the
coherence of these design matrices, some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\gamma$. These simulations are far
from the stringent conditions required by theory to guarantee perfect
recovery, but the performance, while not perfect, may still be good
enough for some purposes.

@@

\begin{figure}
\begin{center}
\subfigure[Forward stepwise with splines]{
\label{fig:spline:fwd}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gauss_glint_p20_n200_lower2p9_upper3p6.pdf}}
\hspace{-15pt}
\subfigure[P-value with splines]{
\label{fig:spline:pval}
\includegraphics[width=0.5\textwidth]{../figs/gamsel_uniform_nsim100_n200_p50_g50_k6_k02_lower10_upper15.pdf}}
\caption{\small \it The left panel shows the power of forward stepwise
for various sparsity levels. The right panel shows our p-value marginally
over each step, with red indicating the global null case and blue
indicating a signal with $k=6$ nonzero groups.}
\label{fig:spline}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Forward stepwise on glinternet]{
\label{fig:glint:fwd}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gauss_glint_p20_n200_lower2p9_upper3p6.pdf}}
\hspace{-15pt}
\subfigure[P-value with glinternet]{
\label{fig:glint:pval}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n200_p40_g40_k6_lower2p5_upper3p9_glint.pdf}}
\caption{\small \it The left panel shows the power of forward stepwise
for various sparsity levels. The right panel shows our p-value marginally
over each step, with red indicating the global null case and blue
indicating a signal with $k=6$ nonzero groups.}
\label{fig:glint}
\end{center}
\end{figure}

>>>>

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some approaches for this problem can be posed as optimization
criteria which stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
where we have written $\{ \beta_s : s = 1, \ldots, steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). Stopping rules based on classical test
statistics have also been used, so it is natural to consider using
the new test statistics of \cite{significance:lasso} or
\cite{tests:adaptive}
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section~\ref{sec:simulations}.

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a subset of the compressed sensing literature \citep{donoho:pursuit, cai:wang:omp} dedicated to forward stepwise (often referred to in that literature as Orthogonal Matching Pursuit or OMP). 
Typical results from these works establish that forward stepwise can exactly select the true model under some stringent conditions involving quantities like the sparsity of the true model and the coherence of the design matrix.
The coherence $\mu(X)$ of a matrix $X$ with columns scaled to have unit 2-norm is defined as
\begin{equation}
  \mu := \mu(X) = \max_{i \neq j} \{ | \innerp{ X_i }{ X_j } | \}
\end{equation}
Recalling that $k$ is the sparsity of $\beta$, it has been shown that if $k < (1/\mu + 1)/2$ and the nonzero coefficients of $\beta$ are sufficiently large then forward stepwise recovers $\beta$ with high probability.
The coherence condition is necessary to guarantee exact recovery \citep{cai:wang:xu:sharp} in the sense that it is possible to construct counterexamples with $k = (1/\mu + 1)/2$.
We refer the reader to \cite{donoho:pursuit, cai:wang:omp} for details.
But for our purposes the conditions required to guarantee exact recovery are usually too stringent.
Simulations show empirically that forward stepwise can work well even when it is not working perfectly, and that it does so under a wide range of conditions.

\begin{figure}
\begin{center}
\subfigure[Categorical design]{
\label{fig:fscat}
\includegraphics[width=0.5\textwidth]{../figs/tpr_default_categorical_nsim100_n100_p400_g200_sizes2-2_k24_lower2_upper4.pdf}}
\hspace{-15pt}
\subfigure[Correlated gaussian design]{
\label{fig:fsgausscorr}
\includegraphics[width=0.5\textwidth]{../figs/tpr_default_gaussian_nsim100_n100_p200_g200_k24_lower1pt2_upper2_corr0pt1.pdf}}
\caption{\small \it The left panel shows the results of the simulation
  using a categorical design consisting of 200 binary factors.
  The right panel shows the results with design matrices of Gaussian
  entries with equi-correlation 0.1 between all columns. Plotted in
  green are the TPP and average model sizes for models chosen by BIC.}
\label{fig:fwdstepsim}
\end{center}
\end{figure}

For various sparsity levels $k$, Algorithm~\ref{algo:fs} was run on
various data sets with \textit{steps} equal to $k$.
In the resulting active set $A_k$, the number of true variables divided by $k$ is the True Positive Proportion (TPP):
\[
\text{TPP}_k = \frac{\# \left\{g: g \in A_k, \beta_g \neq 0\right\}}{k}.
\]
Since we know $k$ in the simulations we can use $\text{TPP}_k$ as a measure of how well forward stepwise is performing. When it is close to 1 we are recovering most of the true variables before including false ones.
Nonzero coefficients had magnitudes
in a range of multiples of $\gamma := \sqrt{2\log(G)/n}$, e.g. in $[1.4
\gamma, 1.6 \gamma]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. After computing the
coherence of these design matrices, some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\gamma$. These simulations are far
from the stringent conditions required by theory to guarantee perfect
recovery, but the performance, while not perfect, may still be good
enough for some purposes. Finally, to serve as a comparison with existing
commonly used model selection procedures, in Figure~\ref{fig:fwdstepsim}
we show in green the TPP and average model sizes for models chosen by
the \textit{step} function in \textit{R} using the BIC criterion.

@@

\begin{figure}
\begin{center}
\subfigure[Forward stepwise with splines]{
\label{fig:spline:fwd}
\includegraphics[width=0.5\textwidth]{../figs/tpr_gamsel_uniform_nsim50_n100_p200_g200_k21_k01_lower5_upper10.pdf}}
\hspace{-15pt}
\subfigure[P-value with splines]{
\label{fig:spline:pval}
\includegraphics[width=0.5\textwidth]{../figs/gamsel_uniform_nsim100_n200_p50_g50_k6_k02_lower10_upper15.pdf}}
\caption{\small \it The left panel shows the power of forward stepwise
for various sparsity levels. The right panel shows our p-value marginally
over each step, with red indicating the global null case and blue
indicating a signal with $k=6$ nonzero groups.}
\label{fig:spline}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfigure[Forward stepwise on glinternet]{
\label{fig:glint:fwd}
\includegraphics[width=0.5\textwidth]{../figs/tpr_glint_categorical_nsim50_n100_p72_g36_sizes2-2_k15_k01_lower3_upper5.pdf}}
\hspace{-15pt}
\subfigure[P-value with glinternet]{
\label{fig:glint:pval}
\includegraphics[width=0.5\textwidth]{../figs/glint_gaussian_nsim100_n100_p30_g30_k6_k02_lower4_upper8.pdf}}
\caption{\small \it The left panel shows the power of forward stepwise
for various sparsity levels. The right panel shows our p-value marginally
over each step, with red indicating the global null case and blue
indicating a signal with $k=6$ nonzero groups.}
\label{fig:glint}
\end{center}
\end{figure}
